\documentclass[12pt]{book}

\title{Think Java}
\author{Allen B. Downey and Chris Mayfield}

\newcommand{\thetitle}{Think Java}
\newcommand{\thesubtitle}{How to Think Like a Computer Scientist}
\newcommand{\theauthors}{Allen B. Downey and Chris Mayfield}
\newcommand{\theversion}{2nd Edition, Version 7.1.0}

%%%% Both LATEX and PLASTEX

\usepackage{graphicx}
\usepackage{setspace}

% automatically index glossary terms
\newcommand{\term}[1]{%
\item[#1:]\index{#1}}

\usepackage{amsmath}
\usepackage{amsthm}

% format end of chapter excercises
\newtheoremstyle{exercise}
  {12pt}        % space above
  {12pt}        % space below
  {}            % body font
  {}            % indent amount
  {\bfseries}   % head font
  {}            % punctuation
  {12pt}        % head space
  {}            % custom head
\theoremstyle{exercise}
\newtheorem{exercise}{Exercise}[chapter]

\newif\ifplastex
\plastexfalse

%%%% PLASTEX ONLY
\ifplastex

\makeindex

\usepackage{localdef}

\usepackage{url}
\renewcommand{\href}[2]{\url{#1}}

\makeatletter
\newcount\anchorcnt
\newcommand*{\Anchor}[1]{%
  \@bsphack%
    \Hy@GlobalStepCount\anchorcnt%
    \edef\@currentHref{anchor.\the\anchorcnt}%
    \Hy@raisedlink{\hyper@anchorstart{\@currentHref}\hyper@anchorend}%
    \M@gettitle{}\label{#1}%
    \@esphack%
}
\makeatother

% code listing environments:
% we don't need these for plastex because they get replaced
% by preprocess.py
%\newenvironment{code}{\begin{verbatim}}{\end{verbatim}}
%\newenvironment{stdout}{\begin{verbatim}}{\end{verbatim}}

% inline syntax formatting
%\newcommand{\java}{\verb}%}
%\newcommand{\java}{\texttt}%}
\newcommand{\java}[1]{{\tt #1}}%{
\newcommand{\textcolor}[1]{\relax}

%%%% LATEX/HTML ONLY
\else

%BEGIN LATEX
\usepackage{comment}
\excludecomment{htmlonly}
\includecomment{latexonly}
%END LATEX

\input{latexonly.tex}

\fi

%%%% END OF PREAMBLE
\begin{document}

\frontmatter

%%%% PLASTEX ONLY
\ifplastex

\maketitle

%%%% LATEX/HTML ONLY
\else

\begin{latexonly}

%--half title-------------------------------------------------
\thispagestyle{empty}

\begin{flushright}
\vspace*{2.0in}

\begin{spacing}{3}
{\huge \thetitle} \\
{\Large \thesubtitle}
\end{spacing}

\vspace{0.25in}

\theversion

\vfill
\end{flushright}

%--verso------------------------------------------------------
\newpage
\thispagestyle{empty}

\quad

%--title page-------------------------------------------------
\newpage
\thispagestyle{empty}

\begin{flushright}
\vspace*{2.0in}

\begin{spacing}{3}
{\huge \thetitle} \\
{\Large \thesubtitle}
\end{spacing}

\vspace{0.25in}

\theversion

\vspace{1in}

{\Large \theauthors}

\vspace{0.5in}

{\Large Green Tea Press}

{\small Needham, Massachusetts}

\vfill
\end{flushright}

%--copyright--------------------------------------------------
\newpage
\thispagestyle{empty}

Copyright \copyright ~2020 \theauthors.

\vspace{0.2in}

\begin{flushleft}
Green Tea Press \\
9 Washburn Ave \\
Needham, MA 02492
\end{flushleft}

Permission is granted to copy, distribute, and/or modify this work under the terms of the Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License, which is available at \url{https://creativecommons.org/licenses/by-nc-sa/4.0/}.

The original form of this book is \LaTeX\ source code.
Compiling this code has the effect of generating a device-independent representation of the book, which can be converted to other formats and printed.

The \LaTeX\ source for this book is available from \url{https://thinkjava.org/} and \url{https://github.com/ChrisMayfield/ThinkJava2}.

%--table of contents------------------------------------------

\cleardoublepage
\setcounter{tocdepth}{1}
\tableofcontents

\end{latexonly}

%--HTML title page--------------------------------------------

\begin{htmlonly}

\vspace{1em}

{\Large \thetitle: \thesubtitle}

{\large \theauthors}

\theversion

\vspace{1em}

Copyright \copyright ~2020 \theauthors.

Permission is granted to copy, distribute, and/or modify this work under the terms of the Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License, which is available at \url{https://creativecommons.org/licenses/by-nc-sa/4.0/}.

\vspace{1em}

\end{htmlonly}

%-------------------------------------------------------------

%%%% END OF THE PART WE SKIP FOR PLASTEX
\fi

\input{ch00.tex}

\mainmatter

%~ \part{Programming Fundamentals}

\input{ch01.tex}

\input{ch02.tex}

\input{ch03.tex}

\input{ch04.tex}

\input{ch05.tex}

\input{ch06.tex}

\input{ch07.tex}

\input{ch08.tex}

%~ \part{Object-Oriented Programming}

\input{ch09.tex}

\input{ch10.tex}

\input{ch11.tex}

\input{ch12.tex}

\input{ch13.tex}

\input{ch14.tex}

\input{ch15.tex}

\input{ch16.tex}

\input{ch17.tex}

\appendix
%\addtocontents{toc}{\protect\newpage}

%~ \part{Appendixes}

%BEGIN LATEX
\renewcommand{\chaptermark}[1]{\markboth{Appendix \thechapter ~~ #1}{}}
%END LATEX

\input{appa.tex}

\input{appb.tex}

\input{appc.tex}

\input{appd.tex}

\backmatter

\printindex

%-------------------------------------------------------------
\ifplastex
\else

\newpage
\thispagestyle{empty}

\vspace*{64pt}

{\bf\huge About the Authors}

\vspace*{40pt}

{\bf Allen Downey} is a professor of computer science at Olin College of Engineering.
He has taught computer science at Wellesley College, Colby College, and UC Berkeley.
He has a PhD in computer science from UC Berkeley and master's and bachelor's degrees from MIT.
Allen is the creator of the best-selling Think series for O'Reilly, which includes {\it Think Python}, {\it Think Complexity}, {\it Think DSP}, and {\it Think Bayes}.

{\bf Chris Mayfield} is an associate professor of computer science at James Madison University, with a research focus on CS education and professional development.
He has a PhD in computer science from Purdue University and bachelor's degrees in CS and German from the University of Utah.

\fi
\end{document}
